---
title: "KATP Gene Correlations (gene-level only, using existing transcript RDS)"
output:
  html_document:
    df_print: paged
---


#Explanation
This script will show KATP channel subunit correlations in our mouse skeletal muscle (AT) data. From this paper:
[Staunton et al](https://link.springer.com/article/10.1186/s13395-021-00283-4)

The data is freely available:
[E-MTAB-10601](https://www.ebi.ac.uk/biostudies/arrayexpress/studies/E-MTAB-10601)

In this script we use only control samples, but young and old.
FULLRUN == TRUE would require the data downloaded and the gtfs installed in an equivilent location to our set up BUT we publish here as the default FULLRUN == FALSE so you can run it locally from Github with just the included rds files, Where preprocessing has already been done.

NB using LFS for the large rds files:
git lfs track "txi_transcripts.rds"

GAI statement: This entire script was passed with Copilot to strip-out the unnecessary chunks we had in our original.

```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

FULLRUN <- FALSE
METADATA_PATH <- "CazAT/metadataCazGemma.csv"
BASE_PATH <- "/Volumes/traveldata/Maisey/CazAT/merged_fastq"
MOUSE_GTF <- "/Volumes/BigDisk/refgenie/genomes/alias/mm10/ensembl_gtf/default/mm10.gtf.gz"

# Cached RDS files used when FULLRUN == FALSE
METADATA_RDS <- "metadata_gene_corr.rds"
TX2GENE_SYMBOLS_RDS <- "tx2gene_mouse_symbols.rds"
TXI_TRANSCRIPTS_RDS <- "txi_transcripts.rds"


# channel subunit -> MGI symbol
KATP_GENES  <- c("Kcnj8", "Kcnj11", "Abcc8", "Abcc9")
KATP_LABELS <- c(Kcnj8 = "Kir6.1", Kcnj11 = "Kir6.2", Abcc8 = "SUR1", Abcc9 = "SUR2")
CORR_METHOD <- "pearson"

```

## Libraries
```{r libraries}
library(tximport)
library(tximeta)
library(GenomicFeatures)
library(AnnotationDbi)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
```

## Cached-object strategy

When `FULLRUN <- FALSE`, this notebook reads only these cached RDS files:

- `metadata_gene_corr.rds`
- `tx2gene_mouse_symbols.rds`
- `txi_transcripts.rds`


## Metadata

```{r metadata}
if (FULLRUN) {
  metadata <- read.csv(METADATA_PATH, stringsAsFactors = FALSE)
  rownames(metadata) <- metadata$sample_id
  metadata$Age <- factor(metadata$Age, levels = c("Young", "Old"))
  saveRDS(metadata, METADATA_RDS)
} else {
  metadata <- readRDS(METADATA_RDS)
}

cat("Loaded", nrow(metadata), "samples\n")
print(table(metadata$Age))
```

## Transcript-to-gene mapping with gene symbols
For gene level analysis:
Only the mapping needed to collapse transcript abundance to gene-level TPM is retained.

```{r tx2gene}
if (FULLRUN) {
  cat("Building transcript-to-gene mapping from GTF...\n")
  txdb <- makeTxDbFromGFF(MOUSE_GTF, format = "gtf")
  k <- keys(txdb, keytype = "TXNAME")
  tx2gene <- AnnotationDbi::select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")
  tx2gene$TXNAME_clean <- sub("\\..*", "", tx2gene$TXNAME)
  tx2gene$GENEID_clean <- sub("\\..*", "", tx2gene$GENEID)

  # Restore your preferred symbol annotation step here if rebuilding from scratch,
  # then save the gene-symbol-annotated object.
  stop("Add/restore your preferred gene-symbol annotation step here, then save to TX2GENE_SYMBOLS_RDS")

  saveRDS(tx2gene, TX2GENE_SYMBOLS_RDS)
} else {
  tx2gene <- readRDS(TX2GENE_SYMBOLS_RDS)
}

cat("Loaded mapping for", nrow(tx2gene), "transcripts\n")
```

## Transcript quantifications (cached RDS)

```{r transcript-import}
if (FULLRUN) {
  quant_files <- sapply(metadata$sample_id, function(sid) {
    sample_base <- sub("_R1\\.fastq\\.gz$", "", sid)
    
    sample_base <- tolower(sample_base)
    quant_dir <- file.path(BASE_PATH, paste0("quants_", sample_base))
    quant_file <- file.path(quant_dir, "quant.sf")
    if (file.exists(quant_file)) {
      quant_file
    } else {
      warning(paste("Missing quant file for:", sid))
      NA_character_
    }
  })

  names(quant_files) <- metadata$sample_id
  valid_samples <- names(quant_files)[!is.na(quant_files)]
  metadata <- metadata[valid_samples, , drop = FALSE]
  saveRDS(metadata, METADATA_RDS)
  quant_files <- quant_files[valid_samples]

  cat("Found", length(quant_files), "quant files\n")

  txi <- tximport(
    quant_files,
    type = "salmon",
    txOut = TRUE,
    ignoreTxVersion = TRUE,
    dropInfReps = FALSE
  )

  saveRDS(txi, TXI_TRANSCRIPTS_RDS)
} else {
  txi <- readRDS(TXI_TRANSCRIPTS_RDS)
}

cat("Imported", nrow(txi$counts), "transcripts across", ncol(txi$counts), "samples\n")
```

## Transcript annotation (for collapsing to genes)

```{r transcript-annotation}
tx_ids <- rownames(txi$counts)
tx_ids_clean <- sub("\\..*", "", tx_ids)

tx_annotation <- data.frame(
  transcript_id = tx_ids,
  transcript_id_clean = tx_ids_clean,
  stringsAsFactors = FALSE
) %>%
  left_join(
    tx2gene %>%
      dplyr::select(TXNAME_clean, GENEID_clean, gene_symbol) %>%
      distinct(),
    by = c("transcript_id_clean" = "TXNAME_clean")
  ) %>%
  dplyr::rename(
    gene_id = GENEID_clean,
    symbol = gene_symbol
  )

cat("Annotated", nrow(tx_annotation), "transcripts\n")
cat("Transcripts with gene symbols:", sum(!is.na(tx_annotation$symbol)), "\n")
cat("Unknown transcripts:", sum(is.na(tx_annotation$symbol)), "\n")
```

## KATP gene-level TPM

```{r katp-gene-tpm}
expr_long <- as.data.frame(t(txi$abundance)) %>%
  mutate(sample_id = rownames(.)) %>%
  pivot_longer(cols = -sample_id, names_to = "transcript_id", values_to = "TPM") %>%
  left_join(metadata %>% dplyr::select(sample_id, Age), by = "sample_id") %>%
  left_join(tx_annotation %>% dplyr::select(transcript_id, gene_id, symbol), by = "transcript_id")

katp_gene_tpm <- expr_long %>%
  dplyr::filter(symbol %in% KATP_GENES) %>%
  group_by(sample_id, Age, gene_id, symbol) %>%
  summarise(TPM = sum(TPM), .groups = "drop") %>%
  mutate(channel = KATP_LABELS[symbol])

print(katp_gene_tpm %>% dplyr::select(sample_id, Age, gene_id, symbol, channel, TPM) %>% head())
```

## KATP expression matrix

```{r katp-matrix}
katp_wide <- katp_gene_tpm %>%
  dplyr::select(sample_id, channel, TPM) %>%
  pivot_wider(names_from = channel, values_from = TPM) %>%
  as.data.frame()

rownames(katp_wide) <- katp_wide$sample_id
katp_wide$sample_id <- NULL

katp_mat <- log2(as.matrix(katp_wide) + 1)

want <- c("Kir6.1", "Kir6.2", "SUR1", "SUR2")
katp_mat <- katp_mat[, intersect(want, colnames(katp_mat)), drop = FALSE]

cat("Subunits in matrix:", paste(colnames(katp_mat), collapse = ", "), "\n")
print(head(katp_mat))
```

## Pairwise gene correlations

```{r correlations}
cor_with_p <- function(mat, method = "pearson") {
  v <- colnames(mat)
  n <- length(v)
  r <- p <- matrix(NA_real_, n, n, dimnames = list(v, v))

  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      ct <- tryCatch(
        cor.test(mat[, i], mat[, j], method = method),
        error = function(e) NULL
      )
      if (!is.null(ct)) {
        r[i, j] <- unname(ct$estimate)
        p[i, j] <- ct$p.value
      }
    }
  }

  list(r = r, p = p)
}

katp_cor <- cor_with_p(katp_mat, method = CORR_METHOD)

cat("Correlation coefficients (", CORR_METHOD, "):\n", sep = "")
print(round(katp_cor$r, 3))

cat("\np-values:\n")
print(signif(katp_cor$p, 3))
```

## Plot correlation matrix

```{r plot-correlation}
star <- function(p) {
  ifelse(
    is.na(p), "",
    ifelse(p < 0.001, "***",
           ifelse(p < 0.01, "**",
                  ifelse(p < 0.05, "*", "")))
  )
}

cor_long <- as.data.frame(as.table(katp_cor$r)) %>%
  dplyr::rename(g1 = Var1, g2 = Var2, r = Freq) %>%
  left_join(
    as.data.frame(as.table(katp_cor$p)) %>%
      dplyr::rename(g1 = Var1, g2 = Var2, p = Freq),
    by = c("g1", "g2")
  ) %>%
  mutate(
    lab = ifelse(
      as.character(g1) == as.character(g2),
      "1",
      paste0(sprintf("%.2f", r), star(p))
    )
  )

ggplot(cor_long, aes(g1, g2, fill = r)) +
  geom_tile(colour = "grey90") +
  geom_text(aes(label = lab), size = 3.5) +
  scale_fill_gradient2(low = "#3B4CC0", mid = "white", high = "#B40426", midpoint = 0.5, limits = c(0, 1)) +
  labs(
    title = "KATP subunit co-expression",
    subtitle = paste0(CORR_METHOD, " on log2(TPM+1), * p<0.05 ** p<0.01 *** p<0.001"),
    x = NULL, y = NULL, fill = "r"
  ) +
  coord_fixed() +
  theme_bw() +
  theme(panel.grid = element_blank())

ggsave("katp_correlation_matrix_gene_level.png", width = 6, height = 5, dpi = 300)
```
