---
title: "KATP Gene Correlations (gene-level only)"
output: html_notebook
---

#Explanation
This script will show correlations between KATP subtypes in our skeletal muscle
data.  Mouse AT young and old from this paper:

And data is freely available here:

Note that FULLRUN <- TRUE requires that the GTFs etc are installed on your
system on the same location: but we also supply RDS files with prepopulated
snapshots so you can run with FULLRUN <- FALSE standalone with just the 
RDS files here within.


```{r setup}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

FULLRUN <- FALSE
METADATA_PATH <- "CazAT/metadataCazGemma.csv"
BASE_PATH <- "/Volumes/traveldata/Maisey/CazAT/merged_fastq"
MOUSE_GTF <- "/Volumes/BigDisk/refgenie/genomes/alias/mm10/ensembl_gtf/default/mm10.gtf.gz"

# channel subunit -> MGI symbol
KATP_GENES  <- c("Kcnj8", "Kcnj11", "Abcc8", "Abcc9")
KATP_LABELS <- c(Kcnj8 = "Kir6.1", Kcnj11 = "Kir6.2", Abcc8 = "SUR1", Abcc9 = "SUR2")
CORR_METHOD <- "pearson"
```

## Libraries

```{r libraries}
library(tximport)
library(GenomicFeatures)
library(AnnotationDbi)
library(dplyr)
library(tidyr)
library(ggplot2)
library(viridis)
```

## Sample metadata

```{r metadata}

metadata <- read.csv(METADATA_PATH, stringsAsFactors = FALSE)
rownames(metadata) <- metadata$sample_id
metadata$Age <- factor(metadata$Age, levels = c("Young", "Old"))

cat("Loaded", nrow(metadata), "samples\n")
print(table(metadata$Age))
```

## Transcript-to-gene mapping

This notebook keeps only the transcript-to-gene mapping required to summarise Salmon quantifications at the **gene** level.

```{r tx2gene}
if (FULLRUN) {
  cat("Building transcript-to-gene mapping from GTF...\n")
  txdb <- makeTxDbFromGFF(MOUSE_GTF, format = "gtf")
  k <- keys(txdb, keytype = "TXNAME")
  tx2gene <- AnnotationDbi::select(txdb, keys = k, columns = "GENEID", keytype = "TXNAME")
  tx2gene$TXNAME_clean <- sub("\\..*", "", tx2gene$TXNAME)
  tx2gene$GENEID_clean <- sub("\\..*", "", tx2gene$GENEID)
  saveRDS(tx2gene, "tx2gene_mouse.rds")
} else {
  tx2gene <- readRDS("tx2gene_mouse.rds")
}

if (FULLRUN) {
  # If you need to rebuild symbol annotation, use your preferred annotation source here.
  # This notebook expects a cached object with gene symbols already added.
  stop("Set FULLRUN <- FALSE unless you also rebuild tx2gene_mouse_symbols.rds")
} else {
  tx2gene <- readRDS("tx2gene_mouse_symbols.rds")
}

cat("Loaded mapping for", nrow(tx2gene), "transcripts\n")
```

## Quant files

```{r quant-files}
quant_files <- sapply(metadata$sample_id, function(sid) {
  sample_base <- sub("_R1\\.fastq\\.gz$", "", sid)
  sample_base <- tolower(sample_base)
  quant_dir <- file.path(BASE_PATH, paste0("quants_", sample_base))
  quant_file <- file.path(quant_dir, "quant.sf")
  if (file.exists(quant_file)) {
    quant_file
  } else {
    warning(paste("Missing quant file for:", sid))
    NA_character_
  }
})

names(quant_files) <- metadata$sample_id
valid_samples <- names(quant_files)[!is.na(quant_files)]
metadata <- metadata[valid_samples, , drop = FALSE]
quant_files <- quant_files[valid_samples]

cat("Found", length(quant_files), "quant files\n")
```

## Import gene-level abundance

```{r gene-import}
# tximport needs transcript and gene columns only
# use the cleaned IDs so ignoreTxVersion = TRUE works consistently

tx2gene_import <- tx2gene %>%
  dplyr::select(TXNAME_clean, GENEID_clean) %>%
  distinct() %>%
  dplyr::rename(tx = TXNAME_clean, gene = GENEID_clean)

if (FULLRUN) {
  txi_gene <- tximport(
    quant_files,
    type = "salmon",
    tx2gene = tx2gene_import,
    ignoreTxVersion = TRUE,
    countsFromAbundance = "no"
  )
  saveRDS(txi_gene, "txi_mouse.rds")
} else {
  txi_gene <- readRDS("txi_mouse.rds")
}

cat("Imported", nrow(txi_gene$counts), "genes across", ncol(txi_gene$counts), "samples\n")
```

## Gene annotation

```{r gene-annotation}
gene_annotation <- tx2gene %>%
  dplyr::select(GENEID_clean, gene_symbol) %>%
  distinct() %>%
  dplyr::rename(gene_id = GENEID_clean, symbol = gene_symbol)

gene_annotation <- gene_annotation[match(rownames(txi_gene$counts), gene_annotation$gene_id), , drop = FALSE]

cat("Genes with symbols:", sum(!is.na(gene_annotation$symbol) & gene_annotation$symbol != ""), "\n")
cat("Genes without symbols:", sum(is.na(gene_annotation$symbol) | gene_annotation$symbol == ""), "\n")
```

## KATP gene-level TPM

```{r katp-gene-tpm}
gene_tpm <- as.data.frame(t(txi_gene$abundance)) %>%
  mutate(sample_id = rownames(.)) %>%
  pivot_longer(cols = -sample_id, names_to = "gene_id", values_to = "TPM") %>%
  left_join(metadata %>% dplyr::select(sample_id, Age), by = "sample_id") %>%
  left_join(gene_annotation, by = "gene_id")

katp_gene_tpm <- gene_tpm %>%
  dplyr::filter(symbol %in% KATP_GENES) %>%
  mutate(channel = KATP_LABELS[symbol])

print(katp_gene_tpm %>% dplyr::select(sample_id, Age, gene_id, symbol, channel, TPM) %>% head())
```

## KATP expression matrix

```{r katp-matrix}
katp_wide <- katp_gene_tpm %>%
  dplyr::select(sample_id, channel, TPM) %>%
  pivot_wider(names_from = channel, values_from = TPM) %>%
  as.data.frame()

rownames(katp_wide) <- katp_wide$sample_id
katp_wide$sample_id <- NULL

katp_mat <- log2(as.matrix(katp_wide) + 1)

# keep a sensible subunit order
want <- c("Kir6.1", "Kir6.2", "SUR1", "SUR2")
katp_mat <- katp_mat[, intersect(want, colnames(katp_mat)), drop = FALSE]

cat("Subunits in matrix:", paste(colnames(katp_mat), collapse = ", "), "\n")
print(head(katp_mat))
```

## Pairwise gene correlations

```{r correlations}
cor_with_p <- function(mat, method = "pearson") {
  v <- colnames(mat)
  n <- length(v)
  r <- p <- matrix(NA_real_, n, n, dimnames = list(v, v))

  for (i in seq_len(n)) {
    for (j in seq_len(n)) {
      ct <- tryCatch(
        cor.test(mat[, i], mat[, j], method = method),
        error = function(e) NULL
      )
      if (!is.null(ct)) {
        r[i, j] <- unname(ct$estimate)
        p[i, j] <- ct$p.value
      }
    }
  }

  list(r = r, p = p)
}

katp_cor <- cor_with_p(katp_mat, method = CORR_METHOD)

cat("Correlation coefficients (", CORR_METHOD, "):\n", sep = "")
print(round(katp_cor$r, 3))

cat("\np-values:\n")
print(signif(katp_cor$p, 3))
```

## Plot correlation matrix

```{r plot-correlation}
star <- function(p) {
  ifelse(
    is.na(p), "",
    ifelse(p < 0.001, "***",
           ifelse(p < 0.01, "**",
                  ifelse(p < 0.05, "*", "")))
  )
}

cor_long <- as.data.frame(as.table(katp_cor$r)) %>%
  dplyr::rename(g1 = Var1, g2 = Var2, r = Freq) %>%
  left_join(
    as.data.frame(as.table(katp_cor$p)) %>%
      dplyr::rename(g1 = Var1, g2 = Var2, p = Freq),
    by = c("g1", "g2")
  ) %>%
  mutate(
    lab = ifelse(
      as.character(g1) == as.character(g2),
      "1",
      paste0(sprintf("%.2f", r), star(p))
    )
  )

ggplot(cor_long, aes(g1, g2, fill = r)) +
  geom_tile(colour = "grey90") +
  geom_text(aes(label = lab), size = 3.5) +
  scale_fill_gradient2(low = "#3B4CC0", mid = "white", high = "#B40426", midpoint = 0, limits = c(-1, 1)) +
  labs(
    title = "KATP subunit co-expression",
    subtitle = paste0(CORR_METHOD, " on log2(TPM+1), * p<0.05 ** p<0.01 *** p<0.001"),
    x = NULL, y = NULL, fill = "r"
  ) +
  coord_fixed() +
  theme_bw() +
  theme(panel.grid = element_blank())

ggsave("katp_correlation_matrix_gene_level.png", width = 6, height = 5, dpi = 300)
```
